\dppTitle

### 2014-05-30 manuscripts/dpp-msbayes/reviews/response.tex

\revcomment{
    Given these caveats, what (in your opinion) is the best usage of this
    approach to phylogeographic analysis?
}
\response{
    ADD!!!
}

\revcomment{
    Reviewer 1 requested that we elaborate about how Robert et al's
    \cite{Robert2011} critique of ABC model choice relates to \dppmsbayes.
}
\response{
    I have expanded:
    \begin{quote}
        I also urge caution when using dpp-msbayes due to the fact that ABC
        methods in general are known to be biased for model choice
        \cite{Robert2011}.
    \end{quote}
    to:
    \begin{quote}
        I also urge caution when using \dppmsbayes due to the lack of
        theoretical validation of Bayesian model choice when the full data are
        replaced by summary statistics that are insufficient for discriminating
        across models under comparison \cite{Robert2011}, which is certainly
        the case here.  Robert et al.\ \cite{Robert2011} demonstrated that ABC
        estimates of model posterior probabilities can be biased when such
        across-model insufficient statistics are used \cite{Robert2011}.
    \end{quote}
    I think this makes it more clear how the findings of \cite{Robert2011} are
    potentially relevant to the posterior probabilities estimated by
    \dppmsbayes and \msb.
}

\section{Reviewer 2's Comments}
\setcounter{commentCounter}{0}

\revcomment{
    The complaint that ``This causes divergence models with more divergence-time
    parameters to integrate over a much greater parameter space with low
    likelihood yet high prior density, resulting in small marginal likelihoods
    relative to models with fewer divergence-time parameters.'' (p.2) is
    essentially stressing the point with Occam's razor argument. Which I deem a
    rather positive feature of Bayesian model choice. A reflection on the
    determination of the prior distribution away from uniform priors thus
    sounds most timely, rather than a criticism of the Bayesian foundations.
}
\response{
    Sorry for the confusion here. This statement was not intened as
    a general criticism of Bayesian foundations. Rather, the intent was to
    highlight why prior choice is very important in Bayesian model choice.
    I have added the following sentences to clarify:
    \begin{quote}
        This serves as a ``natural'' penalty for over-parameterization, which
        is a strength of Bayesian model choice.  However, given the sensitivity
        of marginal likelihoods to the prior, care is needed when selecting
        prior distributions \cite{Lindley1957}.  Selecting distributions that
        are broad or very differently shaped than the likelihood density (e.g.,
        a uniform prior for divergence times) can lead to small marginal
        likelihoods of parameter-rich models even if they are correct.
    \end{quote}
}

\revcomment{
    The current paper claims to have reached a satisfactory prior modelling
    with ``improved robustness, accuracy, and power'' (p.3). If I understand
    correctly, the changes are in replacing a uniform distribution with a Gamma
    or a Dirichlet prior.  Which means introducing a seriously large number of
    hyperparameters into the picture. Having a lot of flexibility in the prior
    also means a lot of variability in the resulting inference. This is a
    general feature of all Bayesian modellings and I am rather surprised by the
    conclusion that things look better by taking this step away from the
    uniform prior. Depending on the problem they could also look worse.
}
\response{
    Both uniform and gamma distributions have two parameters. Thus, the new
    model only has one additional hyperparameter compared to the original
    msBayes model: the concentration parameter on the Dirichlet process prior
    on divergence models. Also, as described in the manuscript, the new method
    offers a uniform prior over divergence models, which would have the same
    number of hyper-parameters as the original method.

    Also, in both msBayes and dpp-msbayes the hyper-parameters are fixed, and
    thus do not contribute variability to the model. While a hiearchical
    approach that places hyper-priors on the various hyper-parameters would be
    interesting to explore for either the msBayes or dpp-msbayes model, it is
    beyond the scope of the current study.

    I have added some text to various places in the manuscript to clarify that
    the hyper-parameters are fixed:
    \begin{quote}
        If the recombination rate \recombinationRate is allowed to be non-zero,
        the prior term $p(\recombinationRate)$ would be added to Equation
        \ref{eq:fullModelCompact}, and the prior would be $\recombinationRate
        \sim Gamma(\gammaShape{\recombinationRate},
        \gammaScale{\recombinationRate})$, \change{where
        \gammaShape{\recombinationRate} and \gammaScale{\recombinationRate} are
        specified by the investigator.}
    \end{quote}

    \begin{quote}
        The priors for the demographic parameters are:
        $\ancestralTheta{} \sim Gamma(\gammaShape{\ancestralTheta{}},
        \gammaScale{\ancestralTheta{}})$,
        $\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
        \gammaScale{\descendantTheta{}{}})$,
        $\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
        \gammaScale{\descendantTheta{}{}})$,
        $\bottleScalar{1}{} \sim Beta(\betaA{\bottleScalar{}{}},
        \betaB{\bottleScalar{}{}})$,
        $\bottleScalar{2}{} \sim Beta(\betaA{\bottleScalar{}{}},
        \betaB{\bottleScalar{}{}})$,
        $\bottleTime{} \sim U(0, 1)$,
        and
        $\migrationRate{} \sim Gamma(\gammaShape{\migrationRate{}},
        \gammaScale{\migrationRate{}})$.
        \change{
        By default, \ancestralTheta{}, \descendantTheta{1}{}, and \descendantTheta{2}{}
        share the same prior (i.e., $\gammaShape{\ancestralTheta{}} =
        \gammaShape{\descendantTheta{}{}}$ and $\gammaScale{\ancestralTheta{}} =
        \gammaScale{\descendantTheta{}{}}$), but a separate gamma-distributed prior can
        be assigned to \ancestralTheta{}.
        Also, the \bottleScalar{1}{}, \bottleScalar{2}{}, and \migrationRate{}
        parameters are optional (i.e., it can be assumed that there is no migration
        between populations of each pair, and that the population size of each
        descendant population has been constant through time).
        For the parameters included in the model, the hyper-parameters of the prior
        distributions are specified by the investigator.}
    \end{quote}

    \begin{quote}
        Each \divTime{} within \divTimeVector is \iid as $\divTime{} \sim
        Gamma(\gammaShape{\divTime{}}, \gammaScale{\divTime{}})$, \change{
        where \gammaShape{\divTime{}} and \gammaScale{\divTime{}} are specified
        by the investigator}.
    \end{quote}
}

@@

\revcomment{
    The ABC part is rather standard, except for the rather strange feature that
    the divergence times are used to construct summary statistics (p.10). I
    find it strange because these times are not observed for the actual data.
    So I must be missing something, which should be spelled out more clearly.
    (And I object to the sentence ``I have introduced a new approximate-Bayesian
    model choice method.'' (p.21) since there is no new ABC technique, simply a
    different form of prior.)
}
\response{
    ADD!!
}

\revcomment{
    I do not understand much the part about (reshuffling) introducing bias as
    detailed on p.11: every approximate method gives a ``biased'' answer in the
    sense this answer is not the true and proper posterior distribution. Using
    a different (re-ordered) vector of statistics provides a different ABC
    outcome, hence a different approximate posterior, for which it seems truly
    impossible to check whether or not it increases the discrepancy from the
    true posterior, compared with the other version. I always find it annoying
    to see the word `bias' used in a vague meaning and esp. within a Bayesian
    setting. All Bayesian methods are biased.
}
\response{
    Regarding the reshuffling of the summary statistics calculated from
    different alignments of sequences, those statistics are simply not
    exchangeable. So, reshuffling them in a manner that is not conistent across
    all simulations and the observed data (as is done in msBayes) is not
    mathematically valid. Also, if elements are exchangeable, their order will
    not affect the likelihood (or the posterior, barring sampling error). Thus,
    if the goal is to approximate the likelihood, I would hope the reshuffling
    would also have little affect on the approximate posterior. The fact that
    \cite{Huang2011} demonstrate that it does strongly affect the posterior
    suggests this reshuffling is not only mathematically incorrect, but also is
    misleading the results.

    ADD!!
}

@@

\revcomment{
    Quoting [43] as concluding that ABC model choice is biased is
    somewhat misleading: the warning therein was that Bayes factors and
    posterior probabilities resulting from those ABC steps could be quite
    unrelated with those based on the whole dataset. That the same authors
    concluded that the proper choice of summary statistics leads to a
    consistent model choice shows ABC model choice is not necessarily
    ``biased''.
}
\response{
    This is a very good point, thank you. I have expanded this section
    to be clearer about the findings of \cite{Robert2011} and how they
    relate to the methods discussed in the paper:
    \begin{quote}
        I also urge caution when using \dppmsbayes due to the lack of
        theoretical validation of Bayesian model choice when the full data are
        replaced by summary statistics that are insufficient for discriminating
        across models under comparison \cite{Robert2011}, which is certainly
        the case here.  Robert et al.\ \cite{Robert2011} demonstrated that ABC
        estimates of model posterior probabilities can be biased when such
        across-model insufficient statistics are used \cite{Robert2011}.
    \end{quote}
}

\revcomment{
    I also fail to understand why the posterior probability of model i should
    be distributed as a uniform (``If the method is unbiased, the points should
    fall near the identity line'') when the data is from model i: this is not a
    p-value but a posterior probability and the posterior probability is not
    the frequentist coverage.
}
\response{
    I agree that the justification of why the identity line should be considered
    ``unbiased'' was too vague and relied entirely on the reference to
    \cite{Huelsenbeck2004}. I have expanded this section to be very
    explicit what is meant by ``bias'' and why it is desirable for the method
    to produce estimates that fall along the identity line:
    \begin{quote}
        Ideally, the estimated posterior probability of the one-divergence model
        should estimate the probability that the one-divergence model is
        correct given the data.  For large numbers of simulation replicates,
        the proportion of the replicates in each bin for which the
        one-divergence model is true will approximate the probability that the
        one-divergence model is correct.  Thus, if the method has the desirable
        behavior such that the estimated posterior probability of
        the one-divergence model is an unbiased estimate of the probability
        that the one-divergence model is correct given the data, the points
        should fall near the identity line.  For example, let us say the method
        estimates a posterior probability of 0.90 for 1000 simulated datasets.
        If the method is accurately estimating the probability that the
        one-divergence model is correct given the data, then the one-divergence
        model should be the true model in approximately 900 of the 1000
        replicates.  Any trend away from the identity line indicates the method
        is biased in the sense that it is not correctly estimating the
        probability that the one-divergence model is correct given the data.
    \end{quote}
}

@@

\revcomment{
    My overall problem with the paper is that, all in all, this is a single
    Monte Carlo study about the performances of some classes of priors and, as
    such, it does not carry enough weight to validate an approach that remains
    highly subjective in the selection of its hyperparameters. I think this
    remains a controlled experiment with simulated data where the true
    parameters are know and the prior is ``true''. This obviously helps in
    getting better performances. In a different and less controlled experiment,
    a new experimenter may (should?) be at a loss as to which hyperparameters
    to chose.
}
\response{
    I agree. However, the goal of the paper is not to demonstrate, for example,
    that gamma probability distributions are generally superior to uniform
    probability distributions for Bayesian statistics. Nor is the goal to
    solve the subjective nature of prior selection in Bayesian statistics.
    My primary goal was to develop an improved comparative method that
    biogeographers can use to investigate the affect of historical events on
    diversification.

    I used extensive simulations and biological data to compare, as fairly as
    possible, the new approach to the current state-of-the-art method for this
    inference problem. I agree the new approach has many weaknesses, including
    the subjective nature of choosing priors when the truth is unknown. However,
    these weaknesses apply to both \dppmsbayes and \msbayes (and to most
    empirical applications of Bayesian statistical methods where extensive
    prior information is lacking). However, the results strongly support
    the overall conclusion that the primary goal was accomplished, and the
    new approach out performs \msb under a range of conditions.
}

\revcomment{
    The conclusion of the paper that ``With improving numerical methods (...),
    advances in Monte Carlo techniques and increasing efficiency of likelihood
    calculations, analyzing rich comparative phylo-geographical models in a
    full-likelihood Bayesian framework is becoming computationally feasible.''
    (p.21) sounds overly optimistic and rather premature. As to this day, I do
    not know of any significant advance in computing the observed likelihood
    for the population genetics models ABC is currently handling. (The SMC
    algorithm of Bouchard-Côté, Sankaraman and Jordan 2012 does not apply to
    Kingman's coalescent, as far as I can judge.) This is certainly a goal
    worth pursuing and borrowing strength from multiple techniques cannot hurt,
    but it remains so far a lofty goal, still beyond our reach... I thus think
    the major message of the paper is to reinforce our earlier calls for
    caution when interpreting the output of an ABC model choice (p.20), or even
    of a regular Bayesian analysis, agreeing that we should aim at seeing ``a
    large amount of posterior uncertainty'' rather than posterior probability
    values close to 0 and 1.
}
\response{
    Full-likelihood methods in this domain are not as lofty of a goal as you
    might think. The likelihood function of the msBayes model is tractable,
    and, in fact, has already been derived and implemented via reversible-jump
    MCMC (see \cite{JeetDiss}; albeit, not readily available yet).
    Also, there are plenty of examples of rich, Kingman-coalescent models
    implemented in full-likelihood Bayesian frameworks.
    A small sample includes \cite{Drummond2005,Drummond2008,Drummond2002,Kuhner1995,Hey2007,Hey2010,Bryant2012,Shapiro2011,Vaughan2014,Heled2010,Lemey2009,Lemey2010,Pybus2012,Baele2012,Minin2008beast,Bloomquist2010}.
}

### 2014-05-30 manuscripts/dpp-msbayes/dpp-msbayes-main.tex

For priors on divergence-time and demographic parameters, \msb uses continuous
uniform probability distributions.
This causes divergence models with more divergence-time parameters to integrate
over a \emph{much} greater parameter space with low likelihood yet high prior
density, resulting in small marginal likelihoods relative to models with fewer
divergence-time parameters \cite{Jeffreys1939,Lindley1957}.
Given that the marginal likelihood of a model weighted by its prior is what
determines its posterior probability, this can cause support for models with
fewer divergence events \cite{Oaks2012,Oaks2014reply}.
Furthermore, \msb uses a discrete uniform prior over the number of divergence
events $1,2,\ldots,\npairs{}$.
Because there are many more possible assignments of populations pairs to
intermediate numbers of divergence events, this imposes a prior on divergence
models that puts most of the prior mass on models with either very few or very
many divergence-time parameters (see Figure 5 of \cite{Oaks2012}; for brevity
I will refer to this prior as ``U-shaped'').
Given that models with many divergence events can have small marginal
likelihoods due to the uniform priors on divergence-time parameters, the
U-shaped prior will effectively create a strong prior preference for models
with very few divergence events.

@@

\begin{linenomath}
To avoid calculating the likelihood terms of Equation \ref{eq:fullModelCompact},
I distill each sequence alignment \alignment{}{} into a vector of insufficient
summary statistics \alignmentSS{}{}, thus replacing the full dataset
$\alignmentVector = (\alignment{1}{1}, \ldots,
    \alignment{\npairs{}}{\nloci{\npairs{}}})$
with vectors of summary statistics for each alignment
$\ssVectorObs = (\alignmentSSObs{1}{1}, \ldots,
    \alignmentSSObs{\npairs{}}{\nloci{\npairs{}}})$.
For each population pair, the means of the summary statistics are calculated
across the \nloci{} loci, reducing the vector to
$\ssVectorObs = (\alignmentSSObs{1}{}, \ldots,
    \alignmentSSObs{\npairs{}}{})$.
This allows us to estimate the approximate joint posterior distribution
{\small
\begin{equation}
    p(\geneTreeVector, \divTimeMapVector, \demographicParamVector, 
    \locusMutationRateScalarVector, \locusRateHetShapeParameter \given
    \ssSpace, \hkyModelVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) =
    \frac{p(\ssSpace \given \geneTreeVector, \hkyModelVector)
        p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
        \locusMutationRateScalarVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)
        p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter)
        p(\locusRateHetShapeParameter)
        p(\divTimeMapVector)
        p(\demographicParamVector)}{
        p(\ssSpace \given \hkyModelVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)},
    \label{eq:approxModelCompact}
\end{equation}
}
where \ssSpace is the multidimensional Euclidean space around the vector of
summary statistics, the radius of which is the tolerance \tol.
The sources of approximation are the insufficiency of the statistics and the
\tol being greater than zero.
I describe the full model in detail before delving into the numerical
method of estimating the approximate model.
\end{linenomath}

@@

\subsubsection{Prior terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The term $p(\locusRateHetShapeParameter)$ is the prior density
function for the shape parameter of the gamma-distributed prior on
rate heterogeneity among loci.
This prior is $\locusRateHetShapeParameter \sim U(1, 20)$.
The prior probability of the vector of locus-specific mutation-rate multipliers
given \locusRateHetShapeParameter then expands out as a product over
the loci
\begin{equation}
    p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter) =
    \prod_{j=1}^{\nlociTotal}
    p(\locusMutationRateScalar{j} \given \locusRateHetShapeParameter),
    \label{eq:locusRateHetPrior}
\end{equation}
where each \locusMutationRateScalar{} is independently and identically
distributed (\iid) as
$\locusMutationRateScalar{} \sim Gamma(\locusRateHetShapeParameter,
1/\locusRateHetShapeParameter)$.
If the recombination rate \recombinationRate is allowed to be
non-zero, the prior term $p(\recombinationRate)$ would be added
to Equation \ref{eq:fullModelCompact}, and the prior would be
$\recombinationRate \sim Gamma(\gammaShape{\recombinationRate},
\gammaScale{\recombinationRate})$.
\end{linenomath}

\begin{linenomath}
The prior term for the demographic parameters, $p(\demographicParamVector)$,
expands out into its components and as a product over the \npairs{}
pairs of populations
\begin{equation}
    p(\demographicParamVector) =
    \prod_{i=1}^{\npairs{}}
    p(\ancestralTheta{i})
    p(\descendantTheta{1}{i})
    p(\descendantTheta{2}{i})
    p(\bottleScalar{1}{i})
    p(\bottleScalar{2}{i})
    p(\bottleTime{i})
    p(\migrationRate{i}).
    \label{eq:demographicPrior}
\end{equation}
The priors for the demographic parameters are:
$\ancestralTheta{} \sim Gamma(\gammaShape{\ancestralTheta{}},
\gammaScale{\ancestralTheta{}})$,
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
$\bottleScalar{1}{} \sim Beta(\betaA{\bottleScalar{}{}},
\betaB{\bottleScalar{}{}})$,
$\bottleScalar{2}{} \sim Beta(\betaA{\bottleScalar{}{}},
\betaB{\bottleScalar{}{}})$,
$\bottleTime{} \sim U(0, 1)$,
and
$\migrationRate{} \sim Gamma(\gammaShape{\migrationRate{}},
\gammaScale{\migrationRate{}})$.
\end{linenomath}

@@

\begin{linenomath}
Each \divTime{} within \divTimeVector is \iid as $\divTime{} \sim
Gamma(\gammaShape{\divTime{}}, \gammaScale{\divTime{}})$.
Thus, given the number of unique divergence-time classes in
\divTimeIndexVector, this determines the probability of prior term
$p(\divTimeVector \given \divTimeIndexVector)$.
The divergence times are in coalescent units relative to the size of
a constant reference population, which I denote \myTheta{C}, that is equal to
the expectation of the prior on the size of the descendant populations
\begin{equation}
    % \myTheta{C} = \mathbb{E}(\dThetas),
    \myTheta{C} = \mathbb{E}(\descendantTheta{}{}),
    \label{eq:thetaC}
\end{equation}
% where \popIndex is the descendant-population index which can be either 1 or 2.
Given the size of the descendant populations are \iid as
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$ and
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
this becomes
\begin{equation}
    \myTheta{C} = \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}.
    \label{eq:thetaCGamma}
\end{equation}
More specifically, the \divTime{} parameters are in units of
$\myTheta{C}/\mutationRate$ generations, which I denote as \globalcoalunit
generations.
Thus, each \divTime{} within \divTimeVector is proportional to time and can be
converted to the number of generations of the reference population, which I
denote \divTime{G_C}, by assuming a mutation rate and multiplying by the
effective size of the reference population
\begin{equation}
    \divTime{G_C} = \divTime \times \frac{\myTheta{C}}{\mutationRate} = \divTime
    \times \frac{
    \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}}
    {\mutationRate}.
    \label{eq:divTimeGenerations}
\end{equation}
Thus, for each of the divergence times in \divTimeVector to be on the same
scale, the relative mutation rates among the pairs of populations are assumed
to be known and fixed according to the user-provided values in
\mutationRateScalarConstantVector.
\end{linenomath}

@@

\subsection{ABC estimation of the posterior of the model}
\subsubsection{Sampling from the prior}
To estimate the approximate posterior of Equation \ref{eq:approxModelCompact},
I use an ABC rejection algorithm.
The first step of this algorithm entails collecting a random sample of
parameter values from the joint prior and their associated summary
statistics.
Each sample is generated by
(1) drawing values of all the model's parameters, which I denote \hpvector{},
from their respective prior distributions;
(2) rescaling the divergence times
$\divTimeMapVector = (\divTimeMap{1}, \ldots, \divTimeMap{\npairs{}})$
from units proportional to time to units proportional to the expected number
of mutations via Equation \ref{eq:divTimeScaled} to get 
$\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots,
\divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$;
(3) simulating gene trees $\geneTreeVector = (\geneTree{1}{1}, \ldots, 
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
for each locus of each population pair by drawing coalescent times from
a multi-population Kingman-coalescent model given the demographic parameters;
(4) simulating sequence alignments 
$\alignmentVector = (\alignment{1}{1}, \ldots, \alignment{\npairs{}}{\nloci{\npairs{}}})$
along the gene trees under the
\hky substitution parameters
$\hkyModelVector = (\hkyModel{1}{1}, \ldots, \hkyModel{\npairs{}}{\nloci{\npairs{}}})$
that have the same number of sequences and sequence lengths as the observed
dataset;
(5) calculating population genetic summary statistics
$\ssVector{} = (\alignmentSS{1}{1}, \ldots, \alignmentSS{\npairs{}}{\nloci{\npairs{}}})$
from the simulated sequence alignments;
% (these are the same statistics
% calculated from the observed alignments, \ssVectorObs{});
and (6) reducing the summary statistics to the means across loci for each
population pair to get
$\ssVector{} = (\alignmentSS{1}{}, \ldots, \alignmentSS{\npairs{}}{})$, which is
the same summary statistic vector estimated from the observed data \ssVectorObs.
After repeating this procedure \numPriorSamples times, I obtain a random
sample of parameter vectors
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$
from the model prior and their associated vectors of summary statistics
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$.

>>>>

For priors on divergence-time and demographic parameters, \msb uses continuous
uniform probability distributions.
\change{
This causes divergence models with more divergence-time parameters to integrate
over a \emph{much} greater parameter space with low likelihood yet high prior
density, which can result in small marginal likelihoods relative to models with
fewer divergence-time parameters \cite{Jeffreys1939,Lindley1957}.
Given that the marginal likelihood of a model weighted by its prior is what
determines its posterior probability, this can cause support for models with
fewer divergence events \cite{Oaks2012,Oaks2014reply}.
This serves as a ``natural'' penalty for over-parameterization, which is a
strength of Bayesian model choice.
However, given the sensitivity of marginal likelihoods to the prior, care is
needed when selecting prior distributions \cite{Lindley1957}.
Selecting distributions that are broad or very differently shaped than the
likelihood density (e.g., a uniform prior for divergence times) can lead to
small marginal likelihoods of parameter-rich models even if they are correct.
}

Furthermore, \msb uses a discrete uniform prior over the number of divergence
events $1,2,\ldots,\npairs{}$.
Because there are many more possible assignments of populations pairs to
intermediate numbers of divergence events, this imposes a prior on divergence
models that puts most of the prior mass on models with either very few or very
many divergence-time parameters (see Figure 5 of \cite{Oaks2012}; for brevity
I will refer to this prior as ``U-shaped'').
Given that models with many divergence events can have small marginal
likelihoods due to the uniform priors on divergence-time parameters, the
U-shaped prior will effectively create a strong prior preference for models
with very few divergence events.

@@

\begin{linenomath}
To avoid calculating the likelihood terms of Equation \ref{eq:fullModelCompact},
I distill each sequence alignment \alignment{}{} into a vector of insufficient
summary statistics \alignmentSS{}{}, thus replacing the full dataset
$\alignmentVector = (\alignment{1}{1}, \ldots,
    \alignment{\npairs{}}{\nloci{\npairs{}}})$
with vectors of summary statistics for each alignment
$\ssVectorObs = (\alignmentSSObs{1}{1}, \ldots,
    \alignmentSSObs{\npairs{}}{\nloci{\npairs{}}})$.
\change{
Optionally, for each population pair, the means of the summary statistics can
be calculated across the \nloci{} loci, and the vector can be further reduced
to $\ssVectorObs = (\alignmentSSObs{1}{}, \ldots,
\alignmentSSObs{\npairs{}}{})$.
Either way, with \ssVectorObs in hand, we can estimate the approximate joint
posterior distribution}
{\small
\begin{equation}
    p(\geneTreeVector, \divTimeMapVector, \demographicParamVector, 
    \locusMutationRateScalarVector, \locusRateHetShapeParameter \given
    \ssSpace, \hkyModelVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) =
    \frac{p(\ssSpace \given \geneTreeVector, \hkyModelVector)
        p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
        \locusMutationRateScalarVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)
        p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter)
        p(\locusRateHetShapeParameter)
        p(\divTimeMapVector)
        p(\demographicParamVector)}{
        p(\ssSpace \given \hkyModelVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)},
    \label{eq:approxModelCompact}
\end{equation}
}
where \ssSpace is the multidimensional Euclidean space around the vector of
summary statistics, the radius of which is the tolerance \tol.
The sources of approximation are the insufficiency of the statistics and the
\tol being greater than zero.
I describe the full model in detail before delving into the numerical
method of estimating the approximate model.
\end{linenomath}

@@

\subsubsection{Prior terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The term $p(\locusRateHetShapeParameter)$ is the prior density
function for the shape parameter of the gamma-distributed prior on
rate heterogeneity among loci.
This prior is $\locusRateHetShapeParameter \sim U(1, 20)$.
The prior probability of the vector of locus-specific mutation-rate multipliers
given \locusRateHetShapeParameter then expands out as a product over
the loci
\begin{equation}
    p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter) =
    \prod_{j=1}^{\nlociTotal}
    p(\locusMutationRateScalar{j} \given \locusRateHetShapeParameter),
    \label{eq:locusRateHetPrior}
\end{equation}
where each \locusMutationRateScalar{} is independently and identically
distributed (\iid) as
$\locusMutationRateScalar{} \sim Gamma(\locusRateHetShapeParameter,
1/\locusRateHetShapeParameter)$.
If the recombination rate \recombinationRate is allowed to be non-zero, the
prior term $p(\recombinationRate)$ would be added to Equation
\ref{eq:fullModelCompact}, and the prior would be $\recombinationRate \sim
Gamma(\gammaShape{\recombinationRate}, \gammaScale{\recombinationRate})$,
\change{where \gammaShape{\recombinationRate} and
\gammaScale{\recombinationRate} are specified by the investigator.}
\end{linenomath}

\begin{linenomath}
The prior term for the demographic parameters, $p(\demographicParamVector)$,
expands out into its components and as a product over the \npairs{}
pairs of populations
\begin{equation}
    p(\demographicParamVector) =
    \prod_{i=1}^{\npairs{}}
    p(\ancestralTheta{i})
    p(\descendantTheta{1}{i})
    p(\descendantTheta{2}{i})
    p(\bottleScalar{1}{i})
    p(\bottleScalar{2}{i})
    p(\bottleTime{i})
    p(\migrationRate{i}).
    \label{eq:demographicPrior}
\end{equation}
The priors for the demographic parameters are:
$\ancestralTheta{} \sim Gamma(\gammaShape{\ancestralTheta{}},
\gammaScale{\ancestralTheta{}})$,
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
$\bottleScalar{1}{} \sim Beta(\betaA{\bottleScalar{}{}},
\betaB{\bottleScalar{}{}})$,
$\bottleScalar{2}{} \sim Beta(\betaA{\bottleScalar{}{}},
\betaB{\bottleScalar{}{}})$,
$\bottleTime{} \sim U(0, 1)$,
and
$\migrationRate{} \sim Gamma(\gammaShape{\migrationRate{}},
\gammaScale{\migrationRate{}})$.
\change{
By default, \ancestralTheta{}, \descendantTheta{1}{}, and \descendantTheta{2}{}
share the same prior (i.e., $\gammaShape{\ancestralTheta{}} =
\gammaShape{\descendantTheta{}{}}$ and $\gammaScale{\ancestralTheta{}} =
\gammaScale{\descendantTheta{}{}}$), but a separate gamma-distributed prior can
be assigned to \ancestralTheta{}.
Also, the \bottleScalar{1}{}, \bottleScalar{2}{}, and \migrationRate{}
parameters are optional (i.e., it can be assumed that there is no migration
between populations of each pair, and that the population size of each
descendant population has been constant through time).
For the parameters included in the model, the hyper-parameters of the prior
distributions are specified by the investigator.}
\end{linenomath}

@@

\begin{linenomath}
Each \divTime{} within \divTimeVector is \iid as $\divTime{} \sim
Gamma(\gammaShape{\divTime{}}, \gammaScale{\divTime{}})$,
\change{
where \gammaShape{\divTime{}} and \gammaScale{\divTime{}} are specified by the
investigator}.
Thus, given the number of unique divergence-time classes in
\divTimeIndexVector, this determines the probability of prior term
$p(\divTimeVector \given \divTimeIndexVector)$.
The divergence times are in coalescent units relative to the size of
a constant reference population, which I denote \myTheta{C}, that is equal to
the expectation of the prior on the size of the descendant populations
\begin{equation}
    % \myTheta{C} = \mathbb{E}(\dThetas),
    \myTheta{C} = \mathbb{E}(\descendantTheta{}{}),
    \label{eq:thetaC}
\end{equation}
% where \popIndex is the descendant-population index which can be either 1 or 2.
Given the size of the descendant populations are \iid as
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$ and
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
this becomes
\begin{equation}
    \myTheta{C} = \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}.
    \label{eq:thetaCGamma}
\end{equation}
More specifically, the \divTime{} parameters are in units of
$\myTheta{C}/\mutationRate$ generations, which I denote as \globalcoalunit
generations.
Thus, each \divTime{} within \divTimeVector is proportional to time and can be
converted to the number of generations of the reference population, which I
denote \divTime{G_C}, by assuming a mutation rate and multiplying by the
effective size of the reference population
\begin{equation}
    \divTime{G_C} = \divTime \times \frac{\myTheta{C}}{\mutationRate} = \divTime
    \times \frac{
    \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}}
    {\mutationRate}.
    \label{eq:divTimeGenerations}
\end{equation}
Thus, for each of the divergence times in \divTimeVector to be on the same
scale, the relative mutation rates among the pairs of populations are assumed
to be known and fixed according to the user-provided values in
\mutationRateScalarConstantVector.
\end{linenomath}

@@

\subsection{ABC estimation of the posterior of the model}
\subsubsection{Sampling from the prior}
To estimate the approximate posterior of Equation \ref{eq:approxModelCompact},
I use an ABC rejection algorithm.
The first step of this algorithm entails collecting a random sample of
parameter values from the joint prior and their associated summary
statistics.
Each sample is generated by
\change{
(1) drawing values of all the model's parameters, which I denote \hpvector{},
from their respective prior distributions;
% (2) rescaling the divergence times
% $\divTimeMapVector = (\divTimeMap{1}, \ldots, \divTimeMap{\npairs{}})$
% from units proportional to time to units proportional to the expected number
% of mutations via Equation \ref{eq:divTimeScaled} to get 
% $\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots,
% \divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$;
(2) simulating gene trees $\geneTreeVector = (\geneTree{1}{1}, \ldots, 
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
for each locus of each population pair by drawing coalescent times from
a multi-population Kingman-coalescent model given the demographic parameters;
(3) simulating sequence alignments 
$\alignmentVector = (\alignment{1}{1}, \ldots, \alignment{\npairs{}}{\nloci{\npairs{}}})$
along the gene trees under the
\hky substitution parameters
$\hkyModelVector = (\hkyModel{1}{1}, \ldots, \hkyModel{\npairs{}}{\nloci{\npairs{}}})$
that have the same number of sequences and sequence lengths as the observed
dataset;
and
(4) calculating population genetic summary statistics
$\ssVector{} = (\alignmentSS{1}{1}, \ldots, \alignmentSS{\npairs{}}{\nloci{\npairs{}}})$
from the simulated sequence alignments;
Optionally, an additional step can be performed to reduce the summary
statistics to the means across loci for each population pair to get
$\ssVector{} = (\alignmentSS{1}{}, \ldots, \alignmentSS{\npairs{}}{})$.
Either way, \ssVector{} contains the same summary statistics as those estimated
from the observed data \ssVectorObs.
}
After repeating this procedure \numPriorSamples times, I obtain a random
sample of parameter vectors
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$
from the model prior and their associated vectors of summary statistics
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$.
